### 2014-06-17 kypchak-morphfest-2014/poster.tex
some work on the poster
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@54509 72bbbca6-d526-0410-a7d9-f06f51895060
progress, some adjustment of the colours, etc.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@54525 72bbbca6-d526-0410-a7d9-f06f51895060

%\definecolor{bordercol}{RGB}{40,40,40}
%\definecolor{bordercol}{RGB}{100,0,0}
\definecolor{bordercol}{RGB}{0,0,0}
%\definecolor{headercol1}{RGB}{186,215,230}
%\definecolor{headercol2}{RGB}{80,80,80}
%\definecolor{headercol1}{RGB}{153,0,0}
%\definecolor{headercol2}{RGB}{153,0,0}
\definecolor{headercolone}{RGB}{6,130,22}
\definecolor{headercoltwo}{RGB}{6,130,22}
%\definecolor{boxcolor}{RGB}{186,215,230}
%\definecolor{boxcolor}{RGB}{110,230,209}
%\definecolor{boxcolor}{RGB}{143,255,236}
%\definecolor{boxcolor}{RGB}{171,255,241}
\definecolor{boxcolor}{RGB}{176,232,223}
\definecolor{headerfontcol}{RGB}{255,255,255}
%\definecolor{boxcolor}{RGB}{186,215,230}
%\definecolor{boxcolor}{RGB}{226,200,200}
%\newcommand{\hilitetwo}[1]{{\addfontfeature{Color=FF000099}#1}}
\newcommand{\hilitetwo}[1]{{\addfontfeature{Color=99333399}#1}}
%\newcommand{\hiliteone}[1]{{\addfontfeature{Color=33993399}#1}}
\newcommand{\hiliteone}[1]{{\addfontfeature{Color=06821699}#1}}
\newcommand{\hilitegrey}[1]{{\addfontfeature{Color=77777799}#1}}

@@

\newcommand{\tags}[1]{\hiliteone{#1}}
\newcommand{\tag}[1]{\hilitegrey{<}\hiliteone{#1}\hilitegrey{>}}

@@

	\scalebox{0.79}{
		\begin{tabular}{lllll}
			\toprule
				 & \textbf{Kyrgyz} & \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
				 & /{\qipa qɯɾʁɯz}/ & /{\qipa qɑzɑq}/ & /{\qipa tɒtɑɾ}/ & /{\qipa qumuq}/ \\
				classification & Eastern & Southern & Northern & Western \\
			\midrule
				\multicolumn{4}{l}{population of speakers} \\
			\midrule
				number & 3M & 8M-12M & 5.4M & 430K \\
				primary & Kyrgyzstan & Kazakhstan & Tatarstan & Dagestan \\
				secondary & China, etc. & China, Mongolia & Bashqortostan & --- \\
			\midrule
				\multicolumn{4}{l}{external influences} \\
			\midrule
				Mongolic & moderate & moderate & light & light \\
				Oghuz & — & — & light & moderate \\
				Persian & heavy & heavy & heavy & heavy \\
				Russian & heavy & heavy & heavy & heavy \\
			\bottomrule
		\end{tabular}
	}
%			\begin{itemize}
%				\item Similar historically to Southern Altay
%				\item Similar by convergence to Kazakh, Uzbek
%			\end{itemize}
%			\item Spoken in
%			\begin{itemize}
%				\item Kyrgyzstan, as co-official language\\
%				\textit{high levels of bilingualism with Russian}
%				\item China, Tajikistan, Uzbekistan
%			\end{itemize}
%			\item Over 3 million native speakers\\(estimate based on data from Ethnologue) %\citep{lewis2009}
%			\item Our transducer based on written Kyrgyz of the former Soviet Union (literary and colloquial standards)
%		\end{itemize}
	}

	\headerbox{Morphological transducers}{name=morphtrans,below=kypchaklgs}{
		\htwo{Morphological transducers}
		\begin{itemize}
			\item Efficient (in speed \& size) models of a language's morphology
			\item Take a surface form, and produce valid lexical form(s)
			\item Take a lexical form, and produce valid surface form(s)
		\end{itemize}
		\scalebox{0.96}{
			`алдым' \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}, \texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		}

@@

		\htwo{Framework: HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
	}

@@

		\headerbox{Categorisation}{name=adjectives,below=exout,column=1}{
		%\headerbox{Morphophonology}{name=morphophon,below=morphtrans}{
			\begin{itemize}
				\item Example:\ morphologically distinct adjective classes (not documented elsewhere)
				\item Other Turkic transducers:\ \texttt{0}-derivation (\hilitetwo{overgenerates})
				\item Our approach:\ categorisation (generates/analyses only correct forms)
			\end{itemize}

		\scalebox{0.63}{
			\begin{tabular}{lllll}
				\toprule
				\textbf{Type} & \textbf{Gloss} & {\small \texttt{\tag{adj}(\tag{comp})}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{subst}}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{advl}}} \\
				\midrule
				\textbf{A1} & \eng{good} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} \\
				\textbf{A2} & \eng{old} & {\qipa иске (искерәк)} & {\qipa иске (искерәк)} & {\qipa — (—)} \\
				\textbf{A3} & \eng{dead} & {\qipa үле (—)} & {\qipa үле (—)} & {\qipa — (—)} \\
				\textbf{A4} & \eng{basic} & {\qipa төп (—)} & {\qipa — (—)} & {\qipa — (—)} \\
				\bottomrule
			\end{tabular}
		}
	}

@@

				\multirow{4}{*}{\textbf{Kyrgyz}} & Wikipedia & 5.3M & 84.51 $\pm$ 2.27 & 3.56 \\
					& News & 4.1M & 91.43 $\pm$ 0.51 & 4.19 \\
					& Religion & 215K & 91.66 $\pm$ 1.81 & 3.99 \\\cmidrule{2-5}
				 (r54474) & Average &  & \hiliteone{89.20} $\pm$ 3.48 &  \\
				\midrule
				\multirow{4}{*}{\textbf{Kazakh}} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 & 2.43 \\
					& News & 3.8M & 92.12 $\pm$ 2.72  & 2.88 \\
					& Religion & 851K & 92.49 $\pm$ 1.66  & 2.63 \\\cmidrule{2-5}
				 (r50547) & Average &  & \hiliteone{90.07} $\pm$ 1.91  & 2.64\\
				\midrule
				\multirow{4}{*}{\textbf{Tatar}} & Wikipedia & 159K & 86.35 $\pm$ 2.17  & 2.24\\
					& News & 5.2M & 89.75 $\pm$ 0.07  & 2.30 \\
					& Religion & 382K & 91.25 $\pm$ 2.55 & 2.24 \\\cmidrule{2-5}
				 (r50260) & Average &  & \hiliteone{89.12} $\pm$ 1.60 & 2.26 \\
				\midrule
				\multirow{3}{*}{\textbf{Kumyk}} & News & 286K &  91.10 $\pm$ 0.86  & 1.53 \\
					& Religion & 227K & 92.47 $\pm$ 1.03  & 1.53 \\\cmidrule{2-5}
				 (r50300) & Average &  & \hiliteone{91.78} $\pm$ 0.94 & 1.53 \\
				\bottomrule
			\end{tabular}\\
		}
\end{center}
			%}
			%%\end{centering}
%		}

@@

		\scalebox{0.92}{
			\begin{tabular}{llll}
				\toprule
					& \textbf{letters} & \textbf{values} & \textbf{examples} \\
				\midrule
					\multirow{2}{*}{\texttt{\textbf{kaz}}} & \multirow{2}{*}{{\qipa и, у, ю}} & {\qipa /əj, əw, jəw/} & {\qipa қ\hiliteone{ию}д\hilitetwo{а}} \eng{chopping down} \\
						&	& {\qipa /ɘj, ɘw, jɘw/} &  {\qipa к\hiliteone{ию}д\hilitetwo{е}} \eng{getting dressed} \\
				\midrule

@@

					\multirow{6}{*}{\texttt{\textbf{kum}}} & \multirow{6}{*}{{\qipa ё, ю}} & \multirow{2}{*}{{\qipa /ø, y/ / C \_}} & {\qipa г\hiliteone{ю}нл\hilitetwo{е}р} \eng{days} \\
						&	&	& {\qipa г\hiliteone{ё}зл\hilitetwo{е}р} \eng{eyes} \\
						&	& \multirow{2}{*}{{\qipa /jø, jy/}} & {\qipa \hiliteone{ю}рекл\hilitetwo{е}р} \eng{hearts}\\
						&	&	& {\qipa \hiliteone{ё}нкюл\hilitetwo{е}р} \eng{darlings} \\
						&	& \multirow{2}{*}{{\qipa /jo, ju/}} & {\qipa \hiliteone{ю}лдузл\hilitetwo{а}р} \eng{stars} \\
						&	&	& {\qipa \hiliteone{ё}лл\hilitetwo{а}р} \eng{roads} \\
				\bottomrule
			\end{tabular}
		}\vspace{0.5em}
			\begin{itemize}
				\item solution:\ hairy \texttt{twol} rules cover majority of examples
				\item unaccounted-for words get a harmony-forcing character
				\item adjust rules for harmony-forcing characters
			\end{itemize}

			\htwo{Loanwords}
				\begin{itemize}
					\item Letters that represent front vowels in native words may represent ``back'' vowels in Russian words
				\end{itemize}
			\scalebox{0.86}{
				\begin{tabular}{lll}
					\toprule
						 & \textbf{native word example} & \textbf{Russian word example} \\
					\midrule
						\texttt{\textbf{kaz}} & {\qipa \hiliteone{е}лд\hilitetwo{і}ң} \eng{country's} & {\qipa Назарба\hiliteone{е}вт\hilitetwo{ы}ң} \eng{Nazarbayev's} \\
						\texttt{\textbf{tat}} & {\qipa гал\hiliteone{и}мн\hilitetwo{ә}р} \eng{scientists} & {\qipa арт\hiliteone{и}стл\hilitetwo{а}р} \eng{artists} \\
						\texttt{\textbf{kum}} & {\qipa с\hiliteone{ё}зл\hilitetwo{е}р} \eng{words} & {\qipa самол\hiliteone{ё}тл\hilitetwo{а}р} \eng{airplanes} \\
					\bottomrule
				\end{tabular}
			}
			\begin{itemize}
				\item solution:\ separate continuation lexicon (messy rules)
			\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{LEXICON N1-RUS\\
					:\%\{☭\%\} N1 ;\\
					\vspace*{-0.5em}\\
					LEXICON Nouns\\
					артист:артист N1-RUS ; ! "artist"\\
					галим:галим N1 ; ! "scientist"}\end{spacing}
					\vspace{0em}
				}\\

			\htwo{Acronyms and numerals}
				\begin{itemize}
					\item \texttt{twol} rules handle phonology for spelt-out words\\
					{\qipa от\hiliteone{ыз}\hilitetwo{да}н} \eng{from thirty}, {\qipa б\hiliteone{ес}\hilitetwo{те}н} \eng{from five}
					\item no phonological triggers available in numerals (etc.)\\
					{\qipa 30-\hilitetwo{да}н} \eng{from 30}, {\qipa 5-\hilitetwo{те}н} \eng{from 5}
					\item solution:\ phonology-triggering characters
				\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{4:4\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "төрт"\\
					5:5\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "бес"\\
					3\%0:3\%0\%\{а\%\}\%\{з\%\}       NUM-DIGIT ; ! "отыз"}\end{spacing}
					\vspace{0em}
				}\\

			\htwo{A resulting messy \texttt{twol} rule}
				\codeex{
					\begin{spacing}{0.4}
					RdYotVow = ё ю Ё Ю ; \\
					AbstractVow = \%\{а\%\} \%\{э\%\} \%\{ү\%\} \%\{о\%\} ; \\
					\\
					"{A} front unrounded harmony" \\
\%\{A\%\}:е <=> \hfill{} [ [ :FrontVow | [ :Vow :ь ] ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}                        [ [ :RdVow :ь ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}     [ [ [ \textbackslash{}[ .\#.\ | :Vow ] ] :RdYotVow ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}      [ :RdYotVow й:0 :RdYotVow :Cns :Cns* ]/[ :0 - й:0 ] \_ ; \\ % ! кююню \\
\hspace*{0.1em}\hfill{} [[\%\{э\%\}:0|\%\{ү\%\}:0] :Cns* ]/[[ :0 - AbstractVow: ] | \%-: ]* \_ ; \\
\hspace*{4em}        except \\
\hspace*{0.1em}\hfill{}       [ :RdYotVow :Cns* \%\{☭\%\}:0 :Cns* ]/[ :0 - \%\{☭\%\}:0 ] \_ ; \\ % ! -ёрлар in Russian words \\
\hspace*{0.1em}\hfill{}                          [ :Cns :р \%\{☭\%\}: \%>: :Cns* ]/:0 \_ ; \\% ! центр and тигр take back vowels \\
\hspace*{0.1em}\hfill{}         [ [ :Vow - :RdYotVow ] :RdYotVow :Cns :Cns* ]/:0 \_ ; \\% ! for some reason the rule above doesn't exclude this condition  e.g., боюнгъа \\
\hspace*{0.1em}\hfill{}                          [ :Vow ]/[ [ :0 - й:0 ] | \%>: ] \_ ;\\
			\vspace{-1em}
					\end{spacing}
				}

@@

		\headerbox{Further information}{name=getting,column=0,below=morphtrans}{
			\begin{itemize}
				\item Part of Apertium Turkic project:\\
				{\small \url{http://wiki.apertium.org/wiki/Apertium\_Turkic}}
				\item Transducers available live at \href{http://turkic.apertium.org/}{\texttt{turkic.apertium.org}}
				\item Source code available from apertium's svn repo
				%info at {\small \url{http://wiki.apertium.org/wiki/apertium-kir}}

				\item Turkic RBMT mailing list (>25 subscribers):\\
				\texttt{apertium-turkic@lists.sourceforge.net}\\
				\vspace{-0.5ex}Feel free to post in any language!\\\vspace{-2.5ex}
				\item See our papers in LREC proceedings\\
				(2012:\ Kyrgyz, 2014:\ Kazakh, Tatar, Kumyk)
				\item And feel free to contact the authors any time!
			\end{itemize}
			%\vspace{2.8ex}
		}

@@

			\htwo{Epenthesis}
				\begin{itemize}
					\item Turn {\texttt \{y\}} into a harmonised high vowel when a vowel doesn't follow the following consonant:\\
					мур\{у\}н → мурун \eng{nose}\\
					мур\{у\}н+\{I\}м → мурдум \eng{my nose}\\
				\end{itemize}
					\vspace{-0.5ex}
\codeex{
{\small \texttt{\%\{y\%\}:Vy <=> [ :LastVowel :Cns* :Cns ]/[:0] \blank{1em}\\
	\vspace{-0.5ex}\hspace{1pt}\hfill{} [ :Cns [ .\#.\ | :Cns ] ]/[ :0 | \%>:]\ ;\\
  \vspace{-0.5ex}\hspace{1ex}where  Vy  in  (  и  ү  и  и  ү  ы  ы  у  у  ы  у  у )\\
  \vspace{-0.5ex}LastVowel  in  (  и  ү  е  э  ө  я  а  ё  о  ы  ю  у )\\
  \vspace{-0.5ex}\hspace{12ex}         matched ;}}
}

@@

			\begin{itemize}
				\item Some combinations of possessive and case morphemes are non-canonical (i.e., not formed simply by concatenation):

			\begin{center}
			\vspace{-0.5em}
{\small
{\qipa		
	\begin{tabular}{llllll}
		\toprule
		case & form & 1SG & 2SG & 3SP  \\
		\midrule
		nom & — & -(I)м & -(I)ң & -(S)I  \\
		acc & -NI & -(I)мдI & -(I)ңдI & -\textbf{(S)Iн}  \\
		gen & -NIн & -(I)мдIн & -(I)ңдIн & -(S)IнIн  \\
		loc & -DA & -(I)мдA & -(I)ңдA & -\textbf{(S)IндA}  \\
		abl & -DAн & -(I)мдAн, & -(I)ндAн, & -\textbf{(S)IнAн}  \\
		~   &     & -\textbf{(I)мAн} & -\textbf{(I)ңAн} \\
		dat & -GA & -\textbf{(I)мA} & -\textbf{(I)ңA} & -\textbf{(S)IнA} \\
		\bottomrule
	\end{tabular}
}}
			\end{center}

>>>>

%\definecolor{bordercol}{RGB}{40,40,40}
%\definecolor{bordercol}{RGB}{100,0,0}
\definecolor{bordercol}{RGB}{0,0,0}
%\definecolor{headercol1}{RGB}{186,215,230}
%\definecolor{headercol2}{RGB}{80,80,80}
%\definecolor{headercol1}{RGB}{153,0,0}
%\definecolor{headercol2}{RGB}{153,0,0}
%\definecolor{headercol}{RGB}{6,130,22}
%\definecolor{headercolone}{RGB}{6,130,22}
\definecolor{headercolone}{RGB}{6,130,22}
\definecolor{headercoltwo}{RGB}{6,130,22}
%\definecolor{boxcolor}{RGB}{186,215,230}
%\definecolor{boxcolor}{RGB}{110,230,209}
%\definecolor{boxcolor}{RGB}{143,255,236}
%\definecolor{boxcolor}{RGB}{171,255,241}
%definecolor{boxcolor}{RGB}{176,232,223}
\definecolor{boxcolor}{RGB}{176,208,223}
\definecolor{headerfontcol}{RGB}{255,255,255}
%\definecolor{boxcolor}{RGB}{186,215,230}
%\definecolor{boxcolor}{RGB}{226,200,200}
%\newcommand{\hilitetwo}[1]{{\addfontfeature{Color=FF000099}#1}}
\newcommand{\hilitetwo}[1]{{\addfontfeature{Color=99333399}#1}}
%\newcommand{\hiliteone}[1]{{\addfontfeature{Color=33993399}#1}}
\newcommand{\hiliteone}[1]{{\addfontfeature{Color=06821699}#1}}
\newcommand{\hilitegrey}[1]{{\addfontfeature{Color=77777799}#1}}

@@

\newcommand{\tags}[1]{\hiliteone{#1}}
\newcommand{\tag}[1]{\hilitegrey{<}\hiliteone{#1}\hilitegrey{>}}
\newcommand{\archiphon}[1]{\hilitegrey{\{}\hiliteone{#1}\hilitegrey{\}}}

@@

	\scalebox{0.78}{
		\begin{tabular}{lllll}
			\toprule
				 & \textbf{Kyrgyz} & \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
				 & /{\qipa qɯɾʁɯz}/ & /{\qipa qɑzɑq}/ & /{\qipa tɒtɑɾ}/ & /{\qipa qumuq}/ \\
				classification & Eastern & Southern & Northern & Western \\
			\midrule
				\multicolumn{4}{l}{population of speakers} \\
			\midrule
				number & 3M & 8M-12M & 5.4M & 430K \\
				primary & Kyrgyzstan & Kazakhstan & Tatarstan & Dagestan \\
				secondary & China, etc. & China, Mongolia & Bashqortostan & --- \\
			\midrule
				\multicolumn{4}{l}{external influences} \\
			\midrule
				Mongolic & moderate & moderate & light & light \\
				Oghuz & — & — & light & moderate \\
				Persian & heavy & heavy & heavy & heavy \\
				Russian & heavy & heavy & heavy & heavy \\
			\bottomrule
		\end{tabular}
	}
%			\begin{itemize}
%				\item Similar historically to Southern Altay
%				\item Similar by convergence to Kazakh, Uzbek
%			\end{itemize}
%			\item Spoken in
%			\begin{itemize}
%				\item Kyrgyzstan, as co-official language\\
%				\textit{high levels of bilingualism with Russian}
%				\item China, Tajikistan, Uzbekistan
%			\end{itemize}
%			\item Over 3 million native speakers\\(estimate based on data from Ethnologue) %\citep{lewis2009}
%			\item Our transducer based on written Kyrgyz of the former Soviet Union (literary and colloquial standards)
%		\end{itemize}
	}

	\headerbox{Morphological transducers}{name=morphtrans,below=kypchaklgs}{
		\htwo{Morphological transducers}
		\begin{itemize}
			\item Efficient (in speed \& size) models of a language's morphology
			\item Take a surface form, and produce valid lexical form(s)
			\item Take a lexical form, and produce valid surface form(s)\\
		%\end{itemize}
		\scalebox{0.98}{
			алдым \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}, \texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		}
		\end{itemize}

@@

		\htwo{Framework: HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
		\htwo{Approach}
		\begin{itemize}
			\item morphotactics implemented in \texttt{lexc}
			\item morphophonology implemented in \texttt{twol}
			\item compiled separately; compose-intersected to single transducer\\
			алдым \hilitetwo{↔} \texttt{ал\hilitegrey{>}\archiphon{D}\archiphon{I}\hilitegrey{>}м} \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}
		\end{itemize}

@@

		\headerbox{Categorisation}{name=adjectives,column=4,below=devel}{
		%\headerbox{Morphophonology}{name=morphophon,below=morphtrans}{
			\begin{itemize}
				\item Example:\ morphologically distinct adjective classes (not documented elsewhere)
				\item Other Turkic transducers:\ \texttt{0}-derivation (\hilitetwo{overgenerates})
				\item Our approach:\ categorisation (generates/analyses only correct forms)
			\end{itemize}

		\scalebox{0.72}{
			\begin{tabular}{lllll}
				\toprule
				\textbf{Type} & \textbf{Gloss} & {\small \texttt{\tag{adj}(\tag{comp})}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{subst}}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{advl}}} \\
				\midrule
				\textbf{A1} & \eng{good} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} \\
				\textbf{A2} & \eng{old} & {\qipa иске (искерәк)} & {\qipa иске (искерәк)} & {\qipa — (—)} \\
				\textbf{A3} & \eng{dead} & {\qipa үле (—)} & {\qipa үле (—)} & {\qipa — (—)} \\
				\textbf{A4} & \eng{basic} & {\qipa төп (—)} & {\qipa — (—)} & {\qipa — (—)} \\
				\bottomrule
			\end{tabular}
		}
	}

@@

				\multirow{4}{*}{\textbf{Kyrgyz}} & Wikipedia & 5.3M & 84.51 $\pm$ 2.27 & 3.56 \\
					& News & 4.1M & 91.43 $\pm$ 0.51 & 4.19 \\
					& Religion & 215K & 91.66 $\pm$ 1.81 & 3.99 \\\cmidrule{2-5}
				 (r54474) & Average &  & \hiliteone{89.20} $\pm$ 3.48 & 3.91 \\
				\midrule
				\multirow{4}{*}{\textbf{Kazakh}} & Wikipedia & 25.6M & 85.61 $\pm$ 1.37 & 2.43 \\
					& News & 3.8M & 92.12 $\pm$ 2.72  & 2.88 \\
					& Religion & 851K & 92.49 $\pm$ 1.66  & 2.63 \\\cmidrule{2-5}
				 (r50547) & Average &  & \hiliteone{90.07} $\pm$ 1.91  & 2.64\\
				\midrule
				\multirow{4}{*}{\textbf{Tatar}} & Wikipedia & 159K & 86.35 $\pm$ 2.17  & 2.24\\
					& News & 5.2M & 89.75 $\pm$ 0.07  & 2.30 \\
					& Religion & 382K & 91.25 $\pm$ 2.55 & 2.24 \\\cmidrule{2-5}
				 (r50260) & Average &  & \hiliteone{89.12} $\pm$ 1.60 & 2.26 \\
				\midrule
				\multirow{3}{*}{\textbf{Kumyk}} & News & 286K &  91.10 $\pm$ 0.86  & 1.53 \\
					& Religion & 227K & 92.47 $\pm$ 1.03  & 1.53 \\\cmidrule{2-5}
				 (r50300) & Average &  & \hiliteone{91.78} $\pm$ 0.94 & 1.53 \\
				\bottomrule
			\end{tabular}\\
		}
\end{center}
			%}
			%%\end{centering}
%		}

@@

		%\scalebox{0.92}{
			\begin{tabular}{llll}
				\toprule
					& \textbf{letters} & \textbf{values} & \textbf{examples} \\
				\midrule
					\multirow{2}{*}{\texttt{\textbf{kaz}}} & \multirow{2}{*}{{\qipa и, у, ю}} & {\qipa /əj, əw, jəw/} & {\qipa қ\hiliteone{ию}д\hilitetwo{а}} \eng{chopping down} \\
						&	& {\qipa /ɘj, ɘw, jɘw/} &  {\qipa к\hiliteone{ию}д\hilitetwo{е}} \eng{getting dressed} \\
				\midrule

@@

					\multirow{6}{*}{\texttt{\textbf{kum}}} & \multirow{6}{*}{{\qipa ё, ю}} & \multirow{2}{*}{{\qipa /ø, y/ / C \_}} & {\qipa г\hiliteone{ю}нл\hilitetwo{е}р} \eng{days} \\
						&	&	& {\qipa г\hiliteone{ё}зл\hilitetwo{е}р} \eng{eyes} \\
						&	& \multirow{2}{*}{{\qipa /jø, jy/}} & {\qipa \hiliteone{ю}рекл\hilitetwo{е}р} \eng{hearts}\\
						&	&	& {\qipa \hiliteone{ё}нкюл\hilitetwo{е}р} \eng{darlings} \\
						&	& \multirow{2}{*}{{\qipa /jo, ju/}} & {\qipa \hiliteone{ю}лдузл\hilitetwo{а}р} \eng{stars} \\
						&	&	& {\qipa \hiliteone{ё}лл\hilitetwo{а}р} \eng{roads} \\
				\bottomrule
			\end{tabular}
		%}
		\vspace{0.5em}
			\begin{itemize}
				\item solution:\ hairy \texttt{twol} rules cover majority of examples
				\item unaccounted-for words get a harmony-forcing character
				\item adjust rules for harmony-forcing characters
			\end{itemize}

			\htwo{Loanwords}
				\begin{itemize}
					\item Letters that represent front vowels in native words may represent ``back'' vowels in Russian words
				\end{itemize}
			\scalebox{0.92}{
				\begin{tabular}{lll}
					\toprule
						 & \textbf{native word example} & \textbf{Russian word example} \\
					\midrule
						\texttt{\textbf{kaz}} & {\qipa \hiliteone{е}лд\hilitetwo{і}ң} \eng{country's} & {\qipa Назарба\hiliteone{е}вт\hilitetwo{ы}ң} \eng{Nazarbayev's} \\
						\texttt{\textbf{tat}} & {\qipa гал\hiliteone{и}мн\hilitetwo{ә}р} \eng{scientists} & {\qipa арт\hiliteone{и}стл\hilitetwo{а}р} \eng{artists} \\
						\texttt{\textbf{kum}} & {\qipa с\hiliteone{ё}зл\hilitetwo{е}р} \eng{words} & {\qipa самол\hiliteone{ё}тл\hilitetwo{а}р} \eng{airplanes} \\
					\bottomrule
				\end{tabular}
			}
			\begin{itemize}
				\item solution:\ separate continuation lexicon (messy rules)
			\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{LEXICON N1-RUS\\
					:\%\{☭\%\} N1 ;\\
					\vspace*{-0.5em}\\
					LEXICON Nouns\\
					артист:артист N1-RUS ; ! "artist"\\
					галим:галим N1 ; ! "scientist"}\end{spacing}
					\vspace{0em}
				}\\

			\htwo{Acronyms and numerals}
				\begin{itemize}
					\item \texttt{twol} rules handle phonology for spelt-out words\\
					{\qipa от\hiliteone{ыз}\hilitetwo{да}н} \eng{from thirty}, {\qipa б\hiliteone{ес}\hilitetwo{те}н} \eng{from five}
					\item no phonological triggers available in numerals
					\item (incorrect phonological triggers in acronyms)\\
					{\qipa 30-\hilitetwo{да}н} \eng{from 30}, {\qipa 5-\hilitetwo{те}н} \eng{from 5}
					\item solution:\ phonology-triggering characters
					\item simplified:\ e.g., \archiphon{с} for all voiceless ostruents
				\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{4:4\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "төрт"\\
					5:5\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "бес"\\
					3\%0:3\%0\%\{а\%\}\%\{з\%\}       NUM-DIGIT ; ! "отыз"}\end{spacing}
					\vspace{0em}
				}\\

			\htwo{A resulting messy \texttt{twol} rule}
				\codeex{
					\begin{spacing}{0.4}
					RdYotVow = ё ю Ё Ю ; \\
					AbstractVow = \%\{а\%\} \%\{э\%\} \%\{ү\%\} \%\{о\%\} ; \\
					\\
					"{A} front unrounded harmony" \\
\%\{A\%\}:е <=> \hfill{} [ [ :FrontVow | [ :Vow :ь ] ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}                        [ [ :RdVow :ь ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}     [ [ [ \textbackslash{}[ .\#.\ | :Vow ] ] :RdYotVow ] :Cns :Cns* ]/:0 \_ ; \\
\hspace*{0.1em}\hfill{}      [ :RdYotVow й:0 :RdYotVow :Cns :Cns* ]/[ :0 - й:0 ] \_ ; \\ % ! кююню \\
\hspace*{0.1em}\hfill{} [[\%\{э\%\}:0|\%\{ү\%\}:0] :Cns* ]/[[ :0 - AbstractVow: ] | \%-: ]* \_ ; \\
\hspace*{4em}        except \\
\hspace*{0.1em}\hfill{}       [ :RdYotVow :Cns* \%\{☭\%\}:0 :Cns* ]/[ :0 - \%\{☭\%\}:0 ] \_ ; \\ % ! -ёрлар in Russian words \\
\hspace*{0.1em}\hfill{}                          [ :Cns :р \%\{☭\%\}: \%>: :Cns* ]/:0 \_ ; \\% ! центр and тигр take back vowels \\
\hspace*{0.1em}\hfill{}         [ [ :Vow - :RdYotVow ] :RdYotVow :Cns :Cns* ]/:0 \_ ; \\% ! for some reason the rule above doesn't exclude this condition  e.g., боюнгъа \\
\hspace*{0.1em}\hfill{}                          [ :Vow ]/[ [ :0 - й:0 ] | \%>: ] \_ ;\end{spacing}
			\vspace{1em}

			\htwo{Related issues}
				\begin{itemize}
					\item Stem-final й combines with following а, о, у to form я, ё, ю
					\item 
				\end{itemize}

@@

		\headerbox{Further information}{name=getting,column=0,below=morphtrans}{
			\begin{itemize}
				\item Part of \textbf{Apertium Turkic} project:\\
				{\small \url{http://wiki.apertium.org/wiki/Apertium\_Turkic}}
				\item Transducers available \textbf{live} at \href{http://turkic.apertium.org/}{\texttt{turkic.apertium.org}}
				\item \textbf{Source code} available from Apertium's svn repo
				%info at {\small \url{http://wiki.apertium.org/wiki/apertium-kir}}

				\item Turkic RBMT \textbf{mailing list} (>25 subscribers):\\
				\texttt{apertium-turkic@lists.sourceforge.net}\\
				\vspace{-0.5ex}Feel free to post in any language!\\\vspace{-2.5ex}
				\item See our papers in LREC proceedings\\
				(2012:\ Kyrgyz, 2014:\ Kazakh, Tatar, Kumyk)
				\item And feel free to contact the authors any time!
			\end{itemize}
			%\vspace{2.8ex}
		}

@@

			\htwo{Epenthesis}
				\begin{itemize}
					\item Turn {\texttt \{y\}} into a harmonised high vowel when a vowel doesn't follow the following consonant:\\
					мур\archiphon{у}н → мурун \eng{nose}\\
					мур\archiphon{у}н+\archiphon{I}м → мурдум \eng{my nose}\\
				\end{itemize}
					\vspace{-0.5ex}
\codeex{
			\begin{spacing}{0.9}
{\small \texttt{\%\{y\%\}:Vy <=> [ :LastVowel :Cns* :Cns ]/[:0] \blank{1em}\\
	\vspace{-0.5ex}\hspace{1pt}\hfill{} [ :Cns [ .\#.\ | :Cns ] ]/[ :0 | \%>:]\ ;\\
  \vspace{-0.5ex}\hspace{1ex}where  Vy  in  (  и  ү  и  и  ү  ы  ы  у  у  ы  у  у )\\
  \vspace{-0.5ex}LastVowel  in  (  и  ү  е  э  ө  я  а  ё  о  ы  ю  у )\\
  \vspace{-0.5ex}\hspace{12ex}         matched ;}}\vspace{-1.5em}
  			\end{spacing}
}

@@

			\begin{itemize}
				\item Some combinations of possessive + case morphemes are unpredicted (i.e., not formed simply by concatenation and application of phonology):
			\end{itemize}
